\lecture{3}{}{Constrained Optimization}
\label{lec:l3}

\begin{intuition}
	Unconstrained optimization is the exception, not the rule. In optimal control, dynamics, actuator bounds, state limits, terminal conditions, and collision-avoidance requirements all appear as constraints. Once constraints are present, the gradient of the objective need no longer vanish at an optimum; instead, it must be balanced against the gradients of the active constraints. This lecture builds the static constrained-optimization theory that later reappears in infinite-dimensional form inside Pontryagin's principle, trajectory transcription, and model predictive control.
\end{intuition}

\subsection{Problem Formulation}
\label{sec:l3-problem-formulation}

\begin{definition}[General constrained optimization problem]\label{def:l3-general-constrained-optimization-problem}
	A finite-dimensional constrained optimization problem has the form
	\begin{align*}
		\min_{x \in \mathbb{R}^n} \quad & f(x) \\
		\text{subject to} \quad & h(x)=0, \\
		                        & g(x)\le 0,
	\end{align*}
	where
	\[
		f:\mathbb{R}^n \to \mathbb{R},
		\qquad
		h:\mathbb{R}^n \to \mathbb{R}^m,
		\qquad
		g:\mathbb{R}^n \to \mathbb{R}^p.
	\]
	Here the equality constraints are written componentwise as
	\[
		h_i(x)=0, \qquad i=1,\dots,m,
	\]
	and the inequality constraints as
	\[
		g_j(x)\le 0, \qquad j=1,\dots,p.
	\]
\end{definition}

\begin{definition}[Feasible set, active set, inactive set]\label{def:l3-feasible-set-active-set-inactive-set}
	The \emph{feasible set} is
	\[
		\mathcal{F} = \{x \in \mathbb{R}^n : h(x)=0,\; g(x)\le 0\}.
	\]
	A feasible point $x \in \mathcal{F}$ is:
	\begin{itemize}
		\item \emph{active} for inequality constraint $j$ if
		      \[
			      g_j(x)=0,
		      \]
		\item \emph{inactive} for inequality constraint $j$ if
		      \[
			      g_j(x)<0.
		      \]
	\end{itemize}
	The active-set index set is
	\[
		\mathcal{A}(x)=\{j \in \{1,\dots,p\}: g_j(x)=0\}.
	\]
\end{definition}

\begin{remark}
	Only active constraints can influence first-order optimality conditions. Inactive constraints are locally irrelevant because a sufficiently small perturbation remains feasible with respect to them.
\end{remark}

\begin{eg}[Two-dimensional geometry]
	\begin{enumerate}[label=(\alph*)]
		\item The set
		      \[
			      \{x \in \mathbb{R}^2 : x_1^2+x_2^2\le 1\}
		      \]
		      is the closed unit disk. At an interior point, no inequality constraint is active. At a boundary point with $x_1^2+x_2^2=1$, the boundary inequality is active.

		\item The set
		      \[
			      \{x \in \mathbb{R}^2 : x_1+x_2=1\}
		      \]
		      is an affine line. Every feasible perturbation must stay tangent to that line.

		\item The set
		      \[
			      \{x \in \mathbb{R}^2 : x_1\ge 0,\; x_2\ge 0,\; x_1+x_2\le 1\}
		      \]
		      is a triangle. At an interior point no inequalities are active; at a point on an edge exactly one is active; at a vertex, two constraints may be active simultaneously.
	\end{enumerate}
\end{eg}

\begin{definition}[Feasible direction]\label{def:l3-feasible-direction}
	Let $x \in \mathcal{F}$. A vector $d \in \mathbb{R}^n$ is a \emph{feasible direction} at $x$ if there exists $\bar t>0$ and a curve $\gamma:[0,\bar t)\to \mathcal{F}$ such that
	\[
		\gamma(0)=x,
		\qquad
		\gamma'(0)=d.
	\]
\end{definition}

\begin{remark}
	For equality constraints, feasible directions are tangent to the constraint manifold. For inequality constraints, feasible directions must point into the feasible region or stay tangent to the active boundary.
\end{remark}

\subsection{Equality-Constrained Optimization and Lagrange Multipliers}
\label{sec:l3-equality-constrained-optimization-and-lagrange-multipliers}

\begin{intuition}
	In unconstrained optimization, stationarity means that the objective cannot decrease along any direction, so the gradient must vanish. Under equality constraints, only \emph{feasible} directions are allowed. The gradient need not vanish; it must only vanish on the tangent space of the feasible manifold. That forces the gradient of the objective to lie in the span of the gradients of the constraints.
\end{intuition}

\begin{definition}[Lagrangian for equality constraints]\label{def:l3-lagrangian-for-equality-constraints}
	For the problem
	\begin{align*}
		\min_{x \in \mathbb{R}^n}\quad & f(x) \\
		\text{subject to}\quad & h(x)=0,
	\end{align*}
	the \emph{Lagrangian} is
	\[
		\mathcal{L}(x,\lambda)=f(x)+\lambda^\top h(x),
		\qquad
		\lambda \in \mathbb{R}^m.
	\]
\end{definition}

\begin{definition}[LICQ for equality constraints]\label{def:l3-licq-for-equality-constraints}
	The \emph{linear independence constraint qualification} (LICQ) at a feasible point $x$ means that the gradients
	\[
		\nabla h_1(x),\dots,\nabla h_m(x)
	\]
	are linearly independent. Equivalently, the Jacobian
	\[
		J_h(x) \in \mathbb{R}^{m \times n}
	\]
	has full row rank $m$.
\end{definition}

\begin{theorem}[Lagrange multiplier theorem]\label{thm:l3-lagrange-multiplier-theorem}
	Assume $f,h_1,\dots,h_m$ are continuously differentiable, and let $x^\star$ be a local minimizer of
	\begin{align*}
		\min_x \quad & f(x) \\
		\text{subject to} \quad & h(x)=0.
	\end{align*}
	If LICQ holds at $x^\star$, then there exists a multiplier $\lambda^\star \in \mathbb{R}^m$ such that
	\[
		\nabla_x \mathcal{L}(x^\star,\lambda^\star)
		=
		\nabla f(x^\star)+J_h(x^\star)^\top \lambda^\star
		=
		0.
	\]
\end{theorem}
\begin{proof}
	Because LICQ holds, the implicit function theorem implies that near $x^\star$ the feasible set
	\[
		\{x : h(x)=0\}
	\]
	is a smooth manifold whose tangent space at $x^\star$ is
	\[
		T_{x^\star}\mathcal{F}
		=
		\{d \in \mathbb{R}^n : J_h(x^\star)d=0\}.
	\]
	Since $x^\star$ is a local minimizer restricted to this manifold, the directional derivative of $f$ must vanish along every tangent direction:
	\[
		\nabla f(x^\star)^\top d = 0
		\qquad \text{for all } d \in T_{x^\star}\mathcal{F}.
	\]
	In other words,
	\[
		\nabla f(x^\star) \in \bigl(T_{x^\star}\mathcal{F}\bigr)^\perp.
	\]
		But
		\[
			T_{x^\star}\mathcal{F}
			=
			\operatorname{null}(J_h(x^\star)),
		\]
		and for every matrix $A$,
		\[
			\operatorname{null}(A)^\perp = \operatorname{range}(A^\top).
		\]
	Hence
	\[
		\nabla f(x^\star) \in \operatorname{range}(J_h(x^\star)^\top).
	\]
	Therefore there exists $\lambda^\star$ such that
	\[
		\nabla f(x^\star)+J_h(x^\star)^\top \lambda^\star = 0.
	\]
\end{proof}

\begin{remark}[Geometric interpretation]
	When there is a single equality constraint $h(x)=0$, the condition becomes
	\[
		\nabla f(x^\star) = -\lambda^\star \nabla h(x^\star).
	\]
	Thus the gradient of the objective is parallel to the gradient of the constraint. Since $\nabla h(x^\star)$ is normal to the feasible surface, the objective gradient has no component along the tangent space. This is the constrained analogue of stationarity.
\end{remark}

\begin{eg}[Minimizing distance to the origin on a line]
	Consider
	\begin{align*}
		\min_{x_1,x_2} \quad & x_1^2+x_2^2 \\
		\text{subject to}\quad & x_1+x_2=1.
	\end{align*}
	The Lagrangian is
	\[
		\mathcal{L}(x_1,x_2,\lambda)=x_1^2+x_2^2+\lambda(x_1+x_2-1).
	\]
	The stationarity equations are
	\[
		2x_1+\lambda=0,
		\qquad
		2x_2+\lambda=0,
	\]
	so
	\[
		x_1=x_2.
	\]
	Using the constraint, we get
	\[
		x_1^\star=x_2^\star=\frac{1}{2}.
	\]
	The multiplier is
	\[
		\lambda^\star=-1.
	\]
	Geometrically, this is the orthogonal projection of the origin onto the affine line $x_1+x_2=1$.
\end{eg}

\begin{remark}[What can go wrong without a constraint qualification]
	Constraint qualifications are not technical decoration; they prevent degenerate constraint geometry. Consider
	\begin{align*}
		\min_{x \in \mathbb{R}} \quad & x \\
		\text{subject to}\quad & x^2=0.
	\end{align*}
	The only feasible point is $x^\star=0$, so it is trivially a global minimizer. But the gradient of the constraint is
	\[
		\nabla h(0)=0,
	\]
	so there is no multiplier $\lambda$ satisfying
	\[
		\nabla f(0)+\lambda \nabla h(0)=1+0=0.
	\]
	LICQ fails, and the Lagrange multiplier theorem correspondingly fails.
\end{remark}

\subsection{Inequality Constraints and KKT Conditions}
\label{sec:l3-inequality-constraints-and-kkt-conditions}

\begin{definition}[Lagrangian for mixed equality and inequality constraints]\label{def:l3-lagrangian-for-mixed-equality-and-inequality-constraints}
	For the problem
	\begin{align*}
		\min_x \quad & f(x) \\
		\text{subject to}\quad & h(x)=0, \\
		                       & g(x)\le 0,
	\end{align*}
	the Lagrangian is
	\[
		\mathcal{L}(x,\lambda,\mu)=f(x)+\lambda^\top h(x)+\mu^\top g(x),
	\]
	where
	\[
		\lambda \in \mathbb{R}^m,
		\qquad
		\mu \in \mathbb{R}^p.
	\]
\end{definition}

\begin{definition}[LICQ for mixed constraints]\label{def:l3-licq-for-mixed-constraints}
	LICQ holds at a feasible point $x$ if the vectors
	\[
		\nabla h_1(x),\dots,\nabla h_m(x),
		\qquad
		\nabla g_j(x)\;\; \text{for } j \in \mathcal{A}(x),
	\]
	are linearly independent.
\end{definition}

\begin{lemma}[Linearized alternative for constrained stationarity]\label{lem:l3-linearized-alternative-for-constrained-stationarity}
	Let
	\[
		a_1,\dots,a_m,b_1,\dots,b_q,c \in \mathbb{R}^n.
	\]
	The following are equivalent:
	\begin{enumerate}[label=(\roman*)]
		\item $c^\top d \ge 0$ for every $d$ satisfying
		      \[
			      a_i^\top d = 0,\quad i=1,\dots,m,
			      \qquad
			      b_j^\top d \le 0,\quad j=1,\dots,q.
		      \]
		\item There exist multipliers $\lambda_1,\dots,\lambda_m \in \mathbb{R}$ and $\mu_1,\dots,\mu_q \ge 0$ such that
		      \[
			      c+\sum_{i=1}^m \lambda_i a_i + \sum_{j=1}^q \mu_j b_j = 0.
		      \]
	\end{enumerate}
\end{lemma}

\begin{remark}
	This is a standard consequence of Farkas' lemma. It is the linear-algebraic mechanism behind the Karush--Kuhn--Tucker theorem.
\end{remark}

\begin{theorem}[KKT necessary conditions]\label{thm:l3-kkt-necessary-conditions}
	Assume $f,h,g$ are continuously differentiable. Let $x^\star$ be a local minimizer of
	\begin{align*}
		\min_x \quad & f(x) \\
		\text{subject to}\quad & h(x)=0, \\
		                       & g(x)\le 0,
	\end{align*}
	and assume LICQ holds at $x^\star$. Then there exist multipliers $\lambda^\star \in \mathbb{R}^m$ and $\mu^\star \in \mathbb{R}^p$ such that:
	\begin{align*}
		\nabla_x \mathcal{L}(x^\star,\lambda^\star,\mu^\star) & = 0
		&& \text{(stationarity)}, \\
		h(x^\star) & = 0,\quad g(x^\star)\le 0
		&& \text{(primal feasibility)}, \\
		\mu^\star & \ge 0
		&& \text{(dual feasibility)}, \\
		\mu_j^\star g_j(x^\star) & = 0,\quad j=1,\dots,p
		&& \text{(complementary slackness)}.
	\end{align*}
\end{theorem}
\begin{proof}
	Let
	\[
		\mathcal{A}^\star = \mathcal{A}(x^\star)
	\]
	denote the active set at the minimizer. Any feasible direction $d$ at $x^\star$ must satisfy the linearized constraints
	\[
		J_h(x^\star)d=0,
		\qquad
		\nabla g_j(x^\star)^\top d \le 0
		\quad \text{for } j \in \mathcal{A}^\star.
	\]
	Since $x^\star$ is a local minimizer, the objective cannot decrease in any feasible direction, so
	\[
		\nabla f(x^\star)^\top d \ge 0
	\]
	for every such $d$.

	Apply the linearized alternative lemma with
	\[
		c=\nabla f(x^\star),
		\qquad
		a_i=\nabla h_i(x^\star),
		\qquad
		b_j=\nabla g_j(x^\star), \;\; j \in \mathcal{A}^\star.
	\]
	We obtain multipliers $\lambda^\star$ and $\mu_j^\star \ge 0$ for $j \in \mathcal{A}^\star$ such that
	\[
		\nabla f(x^\star)+J_h(x^\star)^\top \lambda^\star
		+\sum_{j \in \mathcal{A}^\star}\mu_j^\star \nabla g_j(x^\star)=0.
	\]
	Define
	\[
		\mu_j^\star=0
		\qquad \text{for } j \notin \mathcal{A}^\star.
	\]
	Then stationarity becomes
	\[
		\nabla f(x^\star)+J_h(x^\star)^\top \lambda^\star+J_g(x^\star)^\top \mu^\star=0.
	\]
	Primal feasibility holds because $x^\star$ is feasible. Dual feasibility holds because the multipliers of active inequalities are nonnegative and the inactive ones were set to zero. Finally, if $j$ is inactive then $g_j(x^\star)<0$ and $\mu_j^\star=0$, while if $j$ is active then $g_j(x^\star)=0$. In both cases
	\[
		\mu_j^\star g_j(x^\star)=0.
	\]
\end{proof}

\begin{remark}[Complementary slackness]
	The identity
	\[
		\mu_j^\star g_j(x^\star)=0
	\]
	means that each inequality constraint falls into one of two categories:
	\begin{itemize}
		\item either the constraint is inactive, so it does not influence stationarity and its multiplier vanishes,
		\item or the constraint is active, in which case it may carry a positive multiplier.
	\end{itemize}
	There is no third possibility.
\end{remark}

\begin{eg}[KKT analysis in a simplex]
	Consider
	\begin{align*}
		\min_{x_1,x_2}\quad & (x_1-2)^2+(x_2-1)^2 \\
		\text{subject to}\quad & x_1+x_2-1 \le 0, \\
		                       & -x_1 \le 0, \\
		                       & -x_2 \le 0.
	\end{align*}
	The feasible set is the triangle
	\[
		x_1\ge 0,\quad x_2\ge 0,\quad x_1+x_2\le 1.
	\]
	Since the objective is the squared distance to $(2,1)$, the minimizer should be the closest feasible point, namely
	\[
		x^\star=(1,0).
	\]
	Let
	\[
		g_1(x)=x_1+x_2-1,
		\quad
		g_2(x)=-x_1,
		\quad
		g_3(x)=-x_2.
	\]
	Then
	\[
		\nabla f(x)=
		\begin{bmatrix}
			2(x_1-2) \\
			2(x_2-1)
		\end{bmatrix},
	\]
	so
	\[
		\nabla f(1,0)=
		\begin{bmatrix}
			-2 \\
			-2
		\end{bmatrix}.
	\]
	The active constraints at $(1,0)$ are $g_1$ and $g_3$. Stationarity becomes
	\[
		\begin{bmatrix}
			-2 \\
			-2
		\end{bmatrix}
		+\mu_1
		\begin{bmatrix}
			1 \\
			1
		\end{bmatrix}
		+\mu_2
		\begin{bmatrix}
			-1 \\
			0
		\end{bmatrix}
		+\mu_3
		\begin{bmatrix}
			0 \\
			-1
		\end{bmatrix}
		=0.
	\]
	Complementary slackness forces $\mu_2=0$ because $g_2(1,0)=-1<0$. Hence
	\[
		-2+\mu_1=0,
		\qquad
		-2+\mu_1-\mu_3=0.
	\]
	So
	\[
		\mu_1=2,
		\qquad
		\mu_3=0.
	\]
	All KKT conditions are satisfied, and because the problem is convex, this point is globally optimal.
\end{eg}

\begin{theorem}[KKT sufficient conditions for convex problems]\label{thm:l3-kkt-sufficient-conditions-for-convex-problems}
	Assume:
	\begin{itemize}
		\item $f$ is convex,
		\item each equality constraint is affine,
		\item each inequality function $g_j$ is convex.
	\end{itemize}
	If a feasible point $x^\star$ admits multipliers $(\lambda^\star,\mu^\star)$ satisfying the KKT conditions, then $x^\star$ is a global minimizer.
\end{theorem}
\begin{proof}
	By convexity of $f$,
	\[
		f(x)\ge f(x^\star)+\nabla f(x^\star)^\top(x-x^\star)
		\qquad \text{for all } x.
	\]
	Use stationarity:
	\[
		\nabla f(x^\star)
		=
		-J_h(x^\star)^\top \lambda^\star
		-J_g(x^\star)^\top \mu^\star.
	\]
	Thus
	\[
		f(x)\ge f(x^\star)-\lambda^{\star\top}J_h(x^\star)(x-x^\star)-\mu^{\star\top}J_g(x^\star)(x-x^\star).
	\]
	Since the equality constraints are affine,
	\[
		h(x)=h(x^\star)+J_h(x^\star)(x-x^\star)=J_h(x^\star)(x-x^\star)
	\]
	because $h(x^\star)=0$. Therefore, for every feasible $x$,
	\[
		J_h(x^\star)(x-x^\star)=0.
	\]
	By convexity of each $g_j$,
	\[
		g_j(x)\ge g_j(x^\star)+\nabla g_j(x^\star)^\top(x-x^\star),
	\]
	so
	\[
		\nabla g_j(x^\star)^\top(x-x^\star)\le g_j(x)-g_j(x^\star).
	\]
	Multiply by $\mu_j^\star \ge 0$ and sum:
	\[
		\mu^{\star\top}J_g(x^\star)(x-x^\star)
		\le
		\mu^{\star\top}g(x)-\mu^{\star\top}g(x^\star).
	\]
	If $x$ is feasible, then $g(x)\le 0$. Complementary slackness gives
	\[
		\mu^{\star\top}g(x^\star)=0.
	\]
	Hence
	\[
		\mu^{\star\top}J_g(x^\star)(x-x^\star)\le 0.
	\]
	Substituting into the lower bound for $f(x)$ yields
	\[
		f(x)\ge f(x^\star)
	\]
	for every feasible $x$. Therefore $x^\star$ is a global minimizer.
\end{proof}

\subsection{Second-Order Conditions for Constrained Problems}
\label{sec:l3-second-order-conditions-for-constrained-problems}

\begin{definition}[Critical cone]\label{def:l3-critical-cone}
	Let $(x^\star,\lambda^\star,\mu^\star)$ be a KKT point, and let
	\[
		\mathcal{A}^\star=\mathcal{A}(x^\star).
	\]
	The \emph{critical cone} at $x^\star$ is
	\[
		\mathcal{C}(x^\star,\lambda^\star,\mu^\star)
		=
		\left\{
		d \in \mathbb{R}^n :
		\begin{array}{l}
			J_h(x^\star)d=0, \\
			\nabla g_j(x^\star)^\top d = 0 \quad \text{if } j \in \mathcal{A}^\star \text{ and } \mu_j^\star>0, \\
			\nabla g_j(x^\star)^\top d \le 0 \quad \text{if } j \in \mathcal{A}^\star \text{ and } \mu_j^\star=0, \\
			\nabla f(x^\star)^\top d = 0
		\end{array}
		\right\}.
	\]
\end{definition}

\begin{remark}
	The critical cone consists of first-order feasible directions along which the objective has zero first-order change. These are precisely the directions in which second-order information matters.
\end{remark}

\begin{theorem}[Second-order necessary condition]\label{thm:l3-second-order-necessary-condition}
	Assume $f,h,g$ are twice continuously differentiable, and let $(x^\star,\lambda^\star,\mu^\star)$ satisfy the KKT conditions. If $x^\star$ is a local minimizer, then
	\[
		d^\top \nabla^2_{xx}\mathcal{L}(x^\star,\lambda^\star,\mu^\star)d \ge 0
		\qquad \text{for all } d \in \mathcal{C}(x^\star,\lambda^\star,\mu^\star).
	\]
\end{theorem}
\begin{proof}
	Fix a direction $d$ in the critical cone. Under LICQ one may construct a feasible curve
	\[
		x(t)=x^\star+td+o(t)
	\]
	that respects the active constraints to first order. Because $x^\star$ is a local minimizer and because $d$ lies in the critical cone, the first-order term in the Taylor expansion of the Lagrangian vanishes. Hence the second-order term must be nonnegative:
	\[
		0 \le \frac{d^2}{dt^2}\Big|_{t=0} \mathcal{L}(x(t),\lambda^\star,\mu^\star)
		=
		d^\top \nabla^2_{xx}\mathcal{L}(x^\star,\lambda^\star,\mu^\star)d.
	\]
	This is the constrained analogue of the unconstrained positive-semidefinite Hessian test.
\end{proof}

\begin{theorem}[Second-order sufficient condition]\label{thm:l3-second-order-sufficient-condition}
	Assume $f,h,g$ are twice continuously differentiable, and let $(x^\star,\lambda^\star,\mu^\star)$ satisfy the KKT conditions. If
	\[
		d^\top \nabla^2_{xx}\mathcal{L}(x^\star,\lambda^\star,\mu^\star)d > 0
		\qquad \text{for every nonzero } d \in \mathcal{C}(x^\star,\lambda^\star,\mu^\star),
	\]
	then $x^\star$ is a strict local minimizer.
\end{theorem}
\begin{proof}
	Assume by contradiction that $x^\star$ is not a strict local minimizer. Then there exists a sequence of distinct feasible points $x_k \to x^\star$ with
	\[
		f(x_k)\le f(x^\star).
	\]
	Normalize the directions
	\[
		d_k=\frac{x_k-x^\star}{\norm{x_k-x^\star}}
	\]
	and extract a convergent subsequence $d_k \to d$. Standard first-order feasibility arguments show that $d$ belongs to the critical cone. Expanding the Lagrangian along the sequence and using the KKT conditions shows that the second-order term must be nonpositive along $d$, contradicting the assumed positive definiteness on the critical cone. Therefore $x^\star$ is a strict local minimizer.
\end{proof}

\subsection{Lagrangian Duality}
\label{sec:l3-lagrangian-duality}

\begin{definition}[Dual function]\label{def:l3-dual-function}
	The \emph{Lagrange dual function} is
	\[
		q(\lambda,\mu)
		=
		\inf_{x \in \mathbb{R}^n} \mathcal{L}(x,\lambda,\mu)
		=
		\inf_x \bigl(f(x)+\lambda^\top h(x)+\mu^\top g(x)\bigr),
	\]
	with the convention that the dual variables must satisfy
	\[
		\mu \ge 0.
	\]
\end{definition}

\begin{proposition}[The dual function is concave]\label{prop:l3-the-dual-function-is-concave}
	The function $q(\lambda,\mu)$ is concave in $(\lambda,\mu)$, even if the primal problem is not convex.
\end{proposition}
\begin{proof}
	For each fixed $x$, the map
	\[
		(\lambda,\mu)\mapsto f(x)+\lambda^\top h(x)+\mu^\top g(x)
	\]
	is affine in $(\lambda,\mu)$. The dual function is the pointwise infimum of this family of affine functions. A pointwise infimum of affine functions is concave. Explicitly, for any two dual points $(\lambda_1,\mu_1)$ and $(\lambda_2,\mu_2)$ and any $\theta \in [0,1]$,
	\begin{align*}
		q(\theta(\lambda_1,\mu_1)+(1-\theta)(\lambda_2,\mu_2))
		 & = \inf_x \Bigl[\theta \mathcal{L}(x,\lambda_1,\mu_1)+(1-\theta)\mathcal{L}(x,\lambda_2,\mu_2)\Bigr] \\
		 & \ge \theta \inf_x \mathcal{L}(x,\lambda_1,\mu_1)+(1-\theta)\inf_x \mathcal{L}(x,\lambda_2,\mu_2) \\
		 & = \theta q(\lambda_1,\mu_1)+(1-\theta)q(\lambda_2,\mu_2).
	\end{align*}
\end{proof}

\begin{definition}[Dual problem]\label{def:l3-dual-problem}
	The \emph{dual optimization problem} is
	\begin{align*}
		\max_{\lambda,\mu}\quad & q(\lambda,\mu) \\
		\text{subject to}\quad & \mu \ge 0.
	\end{align*}
\end{definition}

\begin{theorem}[Weak duality]\label{thm:l3-weak-duality}
	Let $x$ be any primal-feasible point and $(\lambda,\mu)$ any dual-feasible point. Then
	\[
		q(\lambda,\mu)\le f(x).
	\]
	In particular,
	\[
		\sup_{\mu \ge 0,\lambda} q(\lambda,\mu)
		\le
		\inf_{x \in \mathcal{F}} f(x).
	\]
\end{theorem}
\begin{proof}
	If $x$ is primal feasible, then
	\[
		h(x)=0,
		\qquad
		g(x)\le 0.
	\]
	If $(\lambda,\mu)$ is dual feasible, then $\mu \ge 0$, so
	\[
		\mu^\top g(x)\le 0.
	\]
	Hence
	\[
		\mathcal{L}(x,\lambda,\mu)
		=
		f(x)+\lambda^\top h(x)+\mu^\top g(x)
		\le
		f(x).
	\]
	By definition of the infimum,
	\[
		q(\lambda,\mu)=\inf_z \mathcal{L}(z,\lambda,\mu)\le \mathcal{L}(x,\lambda,\mu)\le f(x).
	\]
	This proves weak duality.
\end{proof}

\begin{definition}[Duality gap]\label{def:l3-duality-gap}
	The \emph{duality gap} is the difference
	\[
		p^\star-d^\star,
	\]
	where
	\[
		p^\star=\inf_{x \in \mathcal{F}} f(x),
		\qquad
		d^\star=\sup_{\mu \ge 0,\lambda} q(\lambda,\mu).
	\]
	By weak duality, the duality gap is always nonnegative.
\end{definition}

\begin{theorem}[Strong duality under Slater's condition]\label{thm:l3-strong-duality-under-slater-s-condition}
	Consider the convex problem
	\begin{align*}
		\min_x \quad & f(x) \\
		\text{subject to}\quad & h(x)=0, \\
		                       & g(x)\le 0,
	\end{align*}
	where $f$ and each $g_j$ are convex and the equality constraints are affine. Assume there exists a strictly feasible point $\bar x$ such that
	\[
		h(\bar x)=0,
		\qquad
		g_j(\bar x)<0 \quad \text{for all } j.
	\]
	Then
	\[
		p^\star=d^\star.
	\]
\end{theorem}
\begin{proof}[Proof sketch]
	The strict feasibility assumption is Slater's condition. It implies that the convex feasible set has nonempty relative interior, which rules out pathological boundary degeneracies. One then applies a supporting-hyperplane theorem to the convex set of achievable pairs
	\[
		\{(g(x),f(x)) : h(x)=0\}.
	\]
	The supporting hyperplane separating the primal optimum from the strictly infeasible region produces nonnegative multipliers $\mu^\star$ and equality multipliers $\lambda^\star$, yielding a saddle point of the Lagrangian and hence equality of primal and dual optimal values.
\end{proof}

\begin{remark}
	Strong duality is one of the central reasons convex optimization is tractable. When it holds, solving the dual is not merely a lower-bounding exercise; it recovers the exact primal optimum.
\end{remark}

\newpage

\subsection{Quadratic Programming}
\label{sec:l3-quadratic-programming}

\begin{definition}[Quadratic program]\label{def:l3-quadratic-program}
	A \emph{quadratic program} (QP) has the form
	\begin{align*}
		\min_x \quad & \frac{1}{2}x^\top Hx+c^\top x \\
		\text{subject to}\quad & Ax=b, \\
		                       & Gx\le h,
	\end{align*}
	where $H=H^\top$. If $H \succeq 0$, then the objective is convex.
\end{definition}

\begin{remark}
	QPs are the workhorse subproblems of modern optimal control. Linear MPC is a QP solved online at every time step, and SQP methods for nonlinear optimal control repeatedly solve structured QPs.
\end{remark}

\begin{proposition}[Equality-constrained QP KKT system]\label{prop:l3-equality-constrained-qp-kkt-system}
	Consider the equality-constrained QP
	\begin{align*}
		\min_x \quad & \frac{1}{2}x^\top Hx+c^\top x \\
		\text{subject to}\quad & Ax=b,
	\end{align*}
	where $H=H^\top$. Any KKT point $(x^\star,\lambda^\star)$ satisfies the linear system
	\[
		\begin{bmatrix}
			H   & A^\top \\
			A   & 0
		\end{bmatrix}
		\begin{bmatrix}
			x^\star \\
			\lambda^\star
		\end{bmatrix}
		=
		\begin{bmatrix}
			-c \\
			b
		\end{bmatrix}.
	\]
\end{proposition}
\begin{proof}
	The Lagrangian is
	\[
		\mathcal{L}(x,\lambda)=\frac{1}{2}x^\top Hx+c^\top x+\lambda^\top(Ax-b).
	\]
	Stationarity gives
	\[
		Hx+c+A^\top \lambda=0.
	\]
	Primal feasibility gives
	\[
		Ax=b.
	\]
	Writing these equations together yields the stated block linear system.
\end{proof}

\begin{eg}[A small QP solved by hand]
	Consider
	\begin{align*}
		\min_{x_1,x_2}\quad & \frac{1}{2}(x_1^2+2x_2^2)-x_1-2x_2 \\
		\text{subject to}\quad & x_1+x_2=1.
	\end{align*}
	Here
	\[
		H=
		\begin{bmatrix}
			1 & 0 \\
			0 & 2
		\end{bmatrix},
		\qquad
		c=
		\begin{bmatrix}
			-1 \\
			-2
		\end{bmatrix},
		\qquad
		A=
		\begin{bmatrix}
			1 & 1
		\end{bmatrix}.
	\]
	The KKT system is
	\[
		\begin{bmatrix}
			1 & 0 & 1 \\
			0 & 2 & 1 \\
			1 & 1 & 0
		\end{bmatrix}
		\begin{bmatrix}
			x_1 \\
			x_2 \\
			\lambda
		\end{bmatrix}
		=
		\begin{bmatrix}
			1 \\
			2 \\
			1
		\end{bmatrix}.
	\]
	From stationarity,
	\[
		x_1+\lambda=1,
		\qquad
		2x_2+\lambda=2.
	\]
	So
	\[
		x_1=1-\lambda,
		\qquad
		x_2=1-\frac{\lambda}{2}.
	\]
	Using $x_1+x_2=1$,
	\[
		1-\lambda + 1-\frac{\lambda}{2}=1
		\qquad \Longrightarrow \qquad
		\lambda=\frac{2}{3}.
	\]
	Hence
	\[
		x_1^\star=\frac{1}{3},
		\qquad
		x_2^\star=\frac{2}{3}.
	\]
	Since $H\succ 0$, the problem is strictly convex, so this feasible KKT point is the unique global minimizer.
\end{eg}

\subsection{Interior-Point Methods}
\label{sec:l3-interior-point-methods}

\begin{remark}
	KKT conditions describe optimal points, but they do not yet constitute a numerically robust algorithm for inequality-constrained problems. One must still cope with complementarity,
	\[
		\mu_i g_i(x)=0,
	\]
	which is nonsmooth as an equation in the pair $(g_i,\mu_i)$. Interior-point methods replace this singular condition by a smooth perturbed relation and then drive the perturbation to zero.
\end{remark}

\begin{definition}[Logarithmic barrier formulation]\label{def:l3-logarithmic-barrier-formulation}
	Consider the problem
	\begin{align*}
		\min_x \quad & f(x) \\
		\text{subject to}\quad & c(x)=0, \\
		                       & g(x)\le 0,
	\end{align*}
	with $g=(g_1,\dots,g_m)^\top$. For a barrier parameter $\mu>0$, the logarithmic barrier subproblem is
	\begin{align*}
		\min_x \quad & f(x)-\mu\sum_{i=1}^{m}\log\bigl(-g_i(x)\bigr) \\
		\text{subject to}\quad & c(x)=0,
	\end{align*}
	defined only on the strict interior
	\[
		g_i(x)<0,
		\qquad i=1,\dots,m.
	\]
\end{definition}

\begin{remark}[Slack-variable form]
	It is often cleaner to introduce slack variables
	\[
		s \in \mathbb{R}^{m},
		\qquad
		s>0,
	\]
	and rewrite the inequalities as equalities
	\[
		g(x)+s=0.
	\]
	The barrier problem then becomes
	\begin{align*}
		\min_{x,s} \quad & f(x)-\mu\sum_{i=1}^{m}\log(s_i) \\
		\text{subject to}\quad & c(x)=0, \\
		                       & g(x)+s=0.
	\end{align*}
	This formulation is the standard entry point for primal-dual interior-point methods because positivity is handled entirely through the barrier term.
\end{remark}

\begin{definition}[Central path]\label{def:l3-central-path}
	A family of primal-dual points
	\[
		(x(\mu),s(\mu),\lambda(\mu),\nu(\mu)),
		\qquad
		\mu>0,
	\]
	is called the \emph{central path} if it satisfies
	\begin{align*}
		\nabla f(x)+\nabla c(x)^\top \lambda+\nabla g(x)^\top \nu & = 0, \\
		c(x) & = 0, \\
		g(x)+s & = 0, \\
		S\nu & = \mu \mathbf{1}, \\
		s & > 0, \\
		\nu & > 0,
	\end{align*}
	where $S=\operatorname{diag}(s_1,\dots,s_m)$.
\end{definition}

\begin{proposition}[Perturbed KKT system for the barrier problem]\label{prop:l3-perturbed-kkt-system-for-the-barrier-problem}
	The KKT conditions of the slack-variable barrier problem are exactly the central-path equations from \autoref{def:l3-central-path}.
\end{proposition}
\begin{proof}
	Form the Lagrangian
	\[
		\mathcal{L}_{\mu}(x,s,\lambda,\nu)
		=
		f(x)-\mu\sum_{i=1}^{m}\log(s_i)
		+\lambda^\top c(x)
		+\nu^\top(g(x)+s).
	\]
	Stationarity with respect to $x$ gives
	\[
		\nabla f(x)+\nabla c(x)^\top \lambda+\nabla g(x)^\top \nu=0.
	\]
	Stationarity with respect to $s$ gives, componentwise,
	\[
		-\frac{\mu}{s_i}+\nu_i=0,
	\]
	or equivalently
	\[
		s_i\nu_i=\mu.
	\]
	The equality constraints of the barrier problem are
	\[
		c(x)=0,
		\qquad
		g(x)+s=0.
	\]
	Finally, the logarithm forces $s_i>0$, and the relation $\nu_i=\mu/s_i$ then implies $\nu_i>0$. These are exactly the central-path equations.
\end{proof}

\begin{theorem}[Barrier path converges to KKT as $\mu \downarrow 0$]\label{thm:l3-barrier-path-converges-to-kkt}
	Suppose $(x(\mu),s(\mu),\lambda(\mu),\nu(\mu))$ is a bounded family of central-path points for $\mu \downarrow 0$. Then every accumulation point $(x^\star,s^\star,\lambda^\star,\nu^\star)$ satisfies the KKT conditions of the original nonlinear program:
	\begin{align*}
		\nabla f(x^\star)+\nabla c(x^\star)^\top \lambda^\star+\nabla g(x^\star)^\top \nu^\star & = 0, \\
		c(x^\star) & = 0, \\
		g(x^\star)+s^\star & = 0, \\
		s^\star & \ge 0, \\
		\nu^\star & \ge 0, \\
		S^\star \nu^\star & = 0.
	\end{align*}
\end{theorem}
\begin{proof}
	Stationarity and primal feasibility hold for every $\mu$, so they pass to the limit by continuity. Since each $s_i(\mu)>0$ and $\nu_i(\mu)>0$, the limit satisfies
	\[
		s_i^\star \ge 0,
		\qquad
		\nu_i^\star \ge 0.
	\]
	Finally,
	\[
		s_i(\mu)\nu_i(\mu)=\mu
	\]
	for every component, and therefore
	\[
		s_i^\star \nu_i^\star = \lim_{\mu \downarrow 0} \mu = 0.
	\]
	This is complementary slackness. Hence any accumulation point is a KKT point of the original problem.
\end{proof}

\begin{algorithm}[H]
	\DontPrintSemicolon
	\caption{Primal-Dual Interior-Point Method}
\label{alg:l3-primal-dual-interior-point-method}
	\KwIn{Initial strictly feasible or nearly feasible guess $(x,s,\lambda,\nu)$ with $s>0$, $\nu>0$; barrier reduction rule; tolerances}
	\While{KKT residuals and complementarity are not sufficiently small}{
		Choose a target complementarity level $\mu>0$\;
		Linearize the perturbed KKT equations around the current iterate\;
		Solve the resulting primal-dual Newton system for $(\Delta x,\Delta s,\Delta \lambda,\Delta \nu)$\;
		Choose primal and dual step lengths so that
		\[
			s+\alpha_p \Delta s > 0,
			\qquad
			\nu+\alpha_d \Delta \nu > 0
		\]
		remain strictly positive\;
		Update the iterate and reduce $\mu$ according to the centering strategy\;
	}
	\KwOut{Approximate primal-dual KKT point}
\end{algorithm}

\begin{remark}[Why interior-point methods matter for optimal control]
	When direct collocation or multiple shooting transcribes an optimal-control problem into a large sparse NLP, interior-point methods are often the default choice. Their Newton systems respect the banded time structure, and the central-path viewpoint handles large numbers of inequality constraints without explicit active-set combinatorics.
\end{remark}

\subsection{Regularization of KKT Systems}
\label{sec:l3-regularization-of-kkt-systems}

\begin{remark}
	Newton steps for equality-constrained QPs and SQP subproblems require solving saddle-point systems. Even when the original optimization problem is well posed, these linear systems can become singular or can have the wrong inertia if the Hessian model is indefinite on the constraint tangent space.
\end{remark}

\begin{definition}[Regularized KKT matrix]\label{def:l3-regularized-kkt-matrix}
	For an equality-constrained quadratic model
	\begin{align*}
		\min_p \quad & \frac{1}{2}p^\top Wp + g^\top p \\
		\text{subject to}\quad & Ap + c = 0,
	\end{align*}
	a standard regularized KKT system is
	\[
		\begin{bmatrix}
			W+\delta I & A^\top \\
			A          & -\gamma I
		\end{bmatrix}
		\begin{bmatrix}
			p \\
			y
		\end{bmatrix}
		=
		-
		\begin{bmatrix}
			g \\
			c
		\end{bmatrix},
	\]
	where $\delta \ge 0$ and $\gamma \ge 0$ are regularization parameters.
\end{definition}

\begin{proposition}[Quasi-definiteness under regularization]\label{prop:l3-quasi-definiteness-under-regularization}
	If
	\[
		W+\delta I \succ 0
	\]
	and $\gamma>0$, then the regularized KKT matrix in \autoref{def:l3-regularized-kkt-matrix} is nonsingular. More precisely, it has positive definite $(1,1)$ block and negative definite Schur complement in the dual variables.
\end{proposition}
\begin{proof}
	Since $W+\delta I \succ 0$, the primal block is invertible. Taking its Schur complement gives
	\[
		-\gamma I - A(W+\delta I)^{-1}A^\top.
	\]
	The matrix
	\[
		A(W+\delta I)^{-1}A^\top
	\]
	is positive semidefinite, so the Schur complement is strictly negative definite because $\gamma>0$. Therefore both the primal block and its Schur complement are nonsingular, which implies nonsingularity of the full block matrix.
\end{proof}

\begin{remark}[Inertia correction]
	Modern NLP solvers do not merely want an invertible KKT matrix; they want one whose inertia matches that of a local constrained minimization problem. In practice, one increases $\delta$ and, when necessary, $\gamma$ until the factorization has the desired pivot structure. This is called \emph{inertia correction}. It is the constrained analogue of adding Levenberg--Marquardt damping in unconstrained Newton methods.
\end{remark}

\subsection{Merit Functions and Line Search for Constrained Problems}
\label{sec:l3-merit-functions-and-line-search-for-constrained-problems}

\begin{definition}[$\ell_1$ exact-penalty merit function]\label{def:l3-l1-exact-penalty-merit-function}
	For the nonlinear program
	\begin{align*}
		\min_x \quad & f(x) \\
		\text{subject to}\quad & c(x)=0, \\
		                       & g(x)\le 0,
	\end{align*}
	a standard globalization device is the merit function
	\[
		\Phi_\rho(x)
		=
		f(x)
		+
		\rho\|c(x)\|_1
		+
		\rho\|\max\{0,g(x)\}\|_1,
	\]
	where $\rho>0$ is a penalty parameter.
\end{definition}

\begin{remark}[Why a merit function is needed]
	A Newton or SQP step may improve stationarity while worsening feasibility, or vice versa. A merit function converts this vector-valued notion of progress into a scalar quantity suitable for line search. The price is that one must choose $\rho$ large enough that feasibility is not underweighted.
\end{remark}

\begin{proposition}[Merit decrease for linearized feasible SQP steps]\label{prop:l3-merit-decrease-for-linearized-feasible-sqp-steps}
	Consider an SQP step $p_k$ obtained from a QP subproblem whose linearized constraints are satisfied:
	\[
		c(x_k)+\nabla c(x_k)p_k=0,
	\]
	and whose active linearized inequalities are enforced analogously. If the Hessian model is positive definite on the linearized tangent space and $\rho$ is chosen sufficiently large relative to the multiplier estimates, then $p_k$ is a descent direction for the merit function $\Phi_\rho$ from \autoref{def:l3-l1-exact-penalty-merit-function}.
\end{proposition}
\begin{proof}[Proof sketch]
	The directional derivative of $\Phi_\rho$ splits into an objective part and a feasibility part. The QP step is constructed to reduce the quadratic model of the Lagrangian, so on the tangent space it gives negative first-order change in the stationarity part. The exact-penalty term contributes
	\[
		\rho \, D\!\left(\|c(x)\|_1 + \|\max\{0,g(x)\}\|_1\right)[p_k].
	\]
	Because the QP step also reduces the linearized constraint residuals, the feasibility contribution is nonpositive. Choosing $\rho$ larger than the current multiplier magnitudes ensures that any possible objective increase caused by imperfect multiplier estimates is dominated by the penalty on infeasibility. Hence the total directional derivative is negative.
\end{proof}

\begin{algorithm}[H]
	\DontPrintSemicolon
	\caption{SQP with Merit-Function Line Search}
\label{alg:l3-sqp-with-merit-function-line-search}
	\KwIn{Initial guess $x_0$, multiplier estimate $(\lambda_0,\mu_0)$, Hessian model $W_0$, penalty parameter $\rho>0$}
	\For{$k=0,1,2,\dots$ until convergence}{
		Build the QP subproblem by quadratic approximation of the Lagrangian and linearization of the constraints\;
		Regularize the QP KKT system if needed and solve for the step $p_k$ and multiplier update\;
		Choose a step size $\alpha_k \in (0,1]$ by backtracking until
		\[
			\Phi_\rho(x_k+\alpha_k p_k)
			\le
			\Phi_\rho(x_k)+\sigma \alpha_k D\Phi_\rho(x_k)[p_k]
		\]
		holds for some $\sigma \in (0,1)$\;
		Update
		\[
			x_{k+1}=x_k+\alpha_k p_k
		\]
		and refresh the Hessian model and multiplier estimates\;
	}
	\KwOut{Approximate KKT point}
\end{algorithm}

\begin{remark}[Augmented Lagrangian versus pure penalty globalization]
	The $\ell_1$ merit function is simple and effective, but many implementations instead use augmented-Lagrangian merit functions. The reason is conditioning: pure penalties require large $\rho$ values, while multiplier terms let the solver enforce feasibility without forcing the quadratic model to become extremely stiff.
\end{remark}

\begin{remark}
	Lecture 3 gives the static constrained-optimization language used repeatedly later:
	\begin{itemize}
		\item Lagrange multipliers become costates in optimal control,
		\item KKT systems become the algebraic core of QP and NLP solvers,
		\item duality becomes the conceptual bridge between primal trajectory optimization and multiplier-based methods.
	\end{itemize}
	In that sense, constrained optimization is not a side topic; it is the algebraic backbone of modern optimal control computation.
\end{remark}
